---
title: "Getting started with zukomp"
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

```{r setup}
library(zukomp)
```

`zukomp` is a **codec registry that ships with DEFLATE, not a DEFLATE package
with room for extras**. Everything goes through two functions that take raw
vectors and return raw vectors; which codec does the work is a runtime lookup,
so adding one — including one that lives in someone else's package — requires
no change to the public header and no ABI bump.

## Compressing and decompressing

`komp_compress()` names a codec. `level` is optional and codec-specific; the
default is whatever that codec considers balanced.

```{r}
x <- charToRaw(strrep("the quick brown fox ", 500))

z <- komp_compress(x, "gzip")
c(input = length(x), output = length(z))

identical(komp_decompress(z), x)
```

Raw vectors only. Character input is refused, because the encoding to compress
is the caller's decision and not one this package should make quietly.

```{r, error = TRUE}
komp_compress("the quick brown fox")
```

Concatenated gzip members are a single stream, as the format intends:

```{r}
both <- c(komp_compress(charToRaw("hello "), "gzip"),
          komp_compress(charToRaw("world"), "gzip"))
rawToChar(komp_decompress(both))
```

## Detection, and refusing to guess

`komp_decompress()` defaults to `codec = "auto"`, which calls `komp_detect()`.
Detection is a registry property rather than a table of magic bytes kept in one
place, and it refuses to guess: a format with no header to recognise returns
`NA` and has to be named.

```{r}
komp_detect(z)
komp_detect(komp_compress(x, "deflate-raw"))
```

That is deliberate. Guessing wrong on a headerless format does not raise an
error — it returns wrong bytes.

```{r, error = TRUE}
komp_decompress(komp_compress(x, "deflate-raw"))
```

## The registry

`komp_codecs()` is the capability table, and the reason this package can
honestly call itself extensible. Codecs it knows the name of but does not
implement still get a row, so "install the satellite package" is a possible
answer rather than an unrecognised name.

```{r}
komp_codecs()
```

`content_encoding` is the HTTP token for the codec, which is what makes the
table useful to an HTTP client: `zlib` is `deflate` on the wire, and a codec
with no token gets `NA`.

```{r}
komp_codec_available("gzip")
komp_codec_available("zstd")
```

`komp_info()` reports the build itself — ABI version, which codecs are compiled
in, and the vendored sources behind them.

```{r}
komp_info()
```

## Untrusted input

Compressed input from the network can expand enormously. Decompression is
capped at 1 GiB by default, overridable per call with `max_output` or globally
with `options(zukomp.max_output = )`. The cap is enforced by the core stream
driver, so no codec — including a third-party one — can bypass it.

```{r, error = TRUE}
bomb <- komp_compress(raw(10e6), "gzip")
length(bomb)

komp_decompress(bomb, max_output = 1024)
```

`max_ratio` is the second lever, off by default, bounding output against input
rather than in absolute terms:

```{r, error = TRUE}
komp_decompress(bomb, max_ratio = 10)
```

## Errors are structured conditions

Failures are conditions with classes and data, not just messages, so callers
branch on what went wrong rather than matching on text. `zukomp_truncated`,
`zukomp_checksum_error`, `zukomp_invalid_data`, `zukomp_output_limit`,
`zukomp_ratio_limit` and the rest all inherit from `zukomp_error`.

```{r}
e <- tryCatch(komp_decompress(z[1:10]), zukomp_error = function(e) e)

class(e)
conditionMessage(e)
```

Each carries `codec`, `input_bytes`, `output_bytes` and `native_status`:

```{r}
e[c("codec", "input_bytes", "output_bytes", "native_status")]
```

## For package authors

`zukomp` exposes a versioned C ABI through R's registered C-callable mechanism,
so another package can drive the codecs from C without vendoring miniz itself.
A consumer needs both `Imports: zukomp` (to load the DLL) and
`LinkingTo: zukomp` (for the headers), plus a real `importFrom()` directive in
`NAMESPACE` — `Imports:` alone does not load the namespace, and
`R_GetCCallable()` then resolves nothing.

```
# in DESCRIPTION
Imports:    zukomp
LinkingTo:  zukomp
```

```c
#include <zukomp-r.h>

const zukomp_api_v1 *api = zukomp_api();   // resolved lazily, cached
zu_codec c = api->codec_lookup("gzip");
```

A package can also **register a codec of its own** at `ZU_CODEC_VENDOR_BASE`,
and it becomes a first-class citizen of `komp_codecs()` — with zukomp's output
and ratio limits applying to it automatically, because they live in the stream
driver rather than in each codec. `tests/consumer/zukomptest` does exactly
this, as a test.

## Scope

In: `identity`, `deflate-raw`, `zlib`, `gzip`; streaming; detection;
concatenated gzip members; decompression limits; the C ABI.

Not in: ZIP and other archive formats, tar, PNG, encryption, filesystem archive
APIs, and a zlib-compatible ABI. Brotli, zstd, LZ4 and Snappy are planned as
separate satellite packages, so their licences and source size stay out of this
one.
